Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-05 dictionary.tex
Made a home for Motivation and Application
Added the homomorphism R.
I’m starting to think about how to handle the general case in one go,
using etale cohomology rather than group cohomology.

\subsection*{Overview}
Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields Fq into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
Although this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, the result is not perfect.
In this generality, the correct class of sheaves on G are called \emph{quasicharacter sheaves}; when $G$ is connected and of finite type, quasicharacter sheaves on $G$ specialize to character sheaves on $G$.
While every character of $G(\Fq)$ may be converted, canonically, into a quasicharacter sheaf on G from which the character of $G(\Fq)$ may be recovered using the trace of Frobenius on G, generally there exist non-trivial quasicharacter sheaves on G for which the trace of Frobenius is trivial; these are called \emph{invisible quasicharacter sheaves} on $G$.
Invisible quasicharacter sheaves on G are precisely the quasicharacter sheaves on $G$ that are lost in translation under the function-sheaf dictionary.
Assuming only that $G$ is a smooth commutative group scheme over $\Fq$ with finitely generated component group, this paper gives a simple cohomological description of invisible quasicharacter sheaves on $G$. The most interesting case is that of etale commutative group schemes over $\Fq$.

@@

So, what is a quasicharacter sheaf?
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.
The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section,
and the kernel of the Trace of Frobenius is classified by the classes in $\Hh^2(\pi_0({\bar G}),\EEx)$ that are fixed by Frobenius.
In contrast to the case of connected commutative algebraic groups, the Trace of Frobenius is not injective.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
The canonical section of the Trace of Frobenius allows one to convert any quasicharacter of $G(\Fq)$
into a quasicharacter sheaf on $G$.
In this way, quasicharacter sheaves geometrize characters of $G(\Fq)$.

@@

Three examples will illustrate some of the issues that arise when extending the function--sheaf dictionary to smooth commutative group schemes.
First, consider the etale group scheme $\ZZ$ over $\Fq$.
While the $\ell$-adic character $\ZZ \to \EEx$ defined by
$1 \mapsto -1$, has finite order, the character defined by
$1 \mapsto 1+\ell$ does not, although its image is bounded;
on the other hand, the character $\ZZ \to \EEx$ defined by
$1 \mapsto \ell$ has unbounded image.
As we show in the paper, $\ell$-adic local systems on $G$ can only be used to geometrize characters of finite order; quasicharacter sheaves, however, provide the geometrization of all characters of $G(\Fq)$.
Second, consider the etale group scheme $\ZZ \times \ZZ$.
\todo{Build an example of an invisible quasicharacter sheaf here.}
Finally, consider the etale group scheme $\ZZ \times \ZZ$ over $\Fq$ with Galois action determined by $\sigma(n,m) = (m,n)$ when the restriction of $\sigma$ to $\mathbb{F}_{q^2}$ is non-trivial.
Although there are infinitly-many geometric points on this group, it has exactly one $\Fq$-rational point.
In this case, the group carries only one isomorphism class of quasicharacter sheaves -- in particular, it has no invisible quasicharacter sheaves -- so the dictionary is perfect, albeit trivial, in this case.

@@

@@

The category $\QC(G)$ of quasicharacter sheaves on $G$ is a rigid monoidal category
\cite{etingof:09a}*{\S1.10} under the tensor product
$\qcs{L} \otimes \qcs{L'} \ceq (\gqcs{L}\otimes\gqcs{L'}, \mu\otimes\mu', \phi\otimes \phi')$,
with duals given by applying the sheaf hom functor
$\sheafHom(\ - \ ,\EE)$.
The category of quasicharacter sheaves on $G$ is not abelian, so $\QC(G)$
is not a tensor category in the sense of \cite{deligne:02a}*{0.1}.
This rigid monoidal category structure for $\QC(G)$ gives the set $\QCiso{G}$
of isomorphism classes in $\QC(G)$ the structure of a group.
We now begin developing the tools to relate $\QCiso{G}$ to $G(\Fq)^*$.

Let $(\gqcs{L},\phi)$ be a Weil sheaf on $G$. Every $g\in G(\Fq)$
determines a point $\bg$ on $\bG$ fixed by $\Frob{G}$ and
therefore an automorphism $\phi_{\bg}$ of $\gqcs{L}_{\bg}$. Set $\trFrob{(\gqcs{L},\phi)}(g) \ceq \trace(\phi_{\bg}, \gqcs{L}_{\bg})$,
and note that if $(\gqcs{L},\phi) \iso (\gqcs{L'},\phi')$ then $\trFrob{(\gqcs{L},\phi)} = \trFrob{(\gqcs{L'},\phi')}$.
 If $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf
then we will abbreviate $\trFrob{(\gqcs{L},\phi)}$ to $\trFrob{\qcs{L}}$.

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined below.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $g_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that the class that $g_\mathcal{L} = g_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
\QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $g_\qcs{L}$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
f(x+y) = f(x) f(y) \frac{g(\Frob{G}(x),\Frob{G}(y))}{g(x,y)}
\end{equation}
for all $x,y \in G(\bFq)$. Thus, the class $\,^{\Frob{}}g_\qcs{L}$ of $\,^{\Frob{}}g$ coincides
with the class $g_\qcs{L}$ of $g$. This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.

@@

If $G$ is a connected and of finite type,
then $H^2(G(\bFq),\EEx)$ is trivial \todo{Reference?} so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible;
 by contrast, if $G$ is etale then $H^2(G(\bFq),\EEx)$ may be non-trivial and $G$ may admit invisible quasicharacters sheaves.
We will return to the latter case in Section~\ref{sec:etale}.
In the meantime, we end this section with a simple property of $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.

\begin{lemma}
The function \[S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))\] is a group homomorphism.
\end{lemma}

\begin{proof}
We must show:
\[
g_{\qcs{L}\otimes \qcs{L'}} = g_{\qcs{L}} \, g_\qcs{L'}
\]
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.\todo{Under construction.}
\end{proof}

@@

\section{The function--sheaf dictionary for disconnected group schemes}
\label{sec:snake}

>>>>

\subsection*{Overview}
Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields $\Fq$ into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
Although this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, the result is not perfect.
In this generality, the correct class of sheaves on G are called \emph{quasicharacter sheaves}; when $G$ is connected and of finite type, quasicharacter sheaves on $G$ specialize to character sheaves on $G$.
While every character of $G(\Fq)$ may be converted, canonically, into a quasicharacter sheaf on G from which the character of $G(\Fq)$ may be recovered using the trace of Frobenius on G, generally there exist non-trivial quasicharacter sheaves on G for which the trace of Frobenius is trivial; these are called \emph{invisible quasicharacter sheaves} on $G$.
Invisible quasicharacter sheaves on G are precisely the quasicharacter sheaves on $G$ that are lost in translation under the function-sheaf dictionary.
Assuming only that $G$ is a smooth commutative group scheme over $\Fq$ with finitely generated component group, this paper gives a simple cohomological description of invisible quasicharacter sheaves on $G$. The most interesting case is that of etale commutative group schemes over $\Fq$.

@@

So, what is a quasicharacter sheaf?
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.
The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section,
and the kernel of the Trace of Frobenius is classified by $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$.
In particular, in contrast to the case of connected commutative algebraic groups, the Trace of Frobenius may not injective.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
The canonical section of the Trace of Frobenius allows one to convert any quasicharacter of $G(\Fq)$
into a quasicharacter sheaf on $G$.
In this way, quasicharacter sheaves geometrize characters of $G(\Fq)$.

@@

Three examples will illustrate some of the issues that arise when extending the function--sheaf dictionary to etale, commutative group schemes over $\Fq$.
First, consider the etale group scheme $\ZZ$ over $\Fq$.
While the $\ell$-adic character $\ZZ \to \EEx$ defined by
$1 \mapsto -1$, has finite order, the character defined by
$1 \mapsto 1+\ell$ does not, although its image is bounded;
on the other hand, the character $\ZZ \to \EEx$ defined by
$1 \mapsto \ell$ has unbounded image.
As we show in the paper, $\ell$-adic local systems on $G$ can only be used to geometrize characters of finite order; quasicharacter sheaves, however, provide the geometrization of all characters of $G(\Fq)$.
Second, consider the etale group scheme $\ZZ \times \ZZ$.
\todo{Build an example of an invisible quasicharacter sheaf here.}
Finally, consider the etale group scheme $\ZZ \times \ZZ$ over $\Fq$ with Galois action determined by $\sigma(n,m) = (m,n)$ when the restriction of $\sigma$ to $\mathbb{F}_{q^2}$ is non-trivial.
Although there are infinitly-many geometric points on this group, it has exactly one $\Fq$-rational point.
In this case, the group carries only one isomorphism class of quasicharacter sheaves -- in particular, it has no invisible quasicharacter sheaves -- so the dictionary is perfect, albeit trivial, in this case.

@@

The category $\QC(G)$ of quasicharacter sheaves on $G$ is a rigid monoidal category
\cite{etingof:09a}*{\S1.10} under the tensor product
$\qcs{L} \otimes \qcs{L'} \ceq (\gqcs{L}\otimes\gqcs{L'}, \mu\otimes\mu', \phi\otimes \phi')$,
with duals given by applying the sheaf hom functor
$\sheafHom(\ - \ ,\EE)$.
The category of quasicharacter sheaves on $G$ is not abelian, so $\QC(G)$
is not a tensor category in the sense of \cite{deligne:02a}*{0.1}.
This rigid monoidal category structure for $\QC(G)$ gives the set $\QCiso{G}$
of isomorphism classes in $\QC(G)$ the structure of a group.

@@

We now begin developing the tools to relate $\QCiso{G}$ to $G(\Fq)^*$.
%
Let $(\gqcs{L},\phi)$ be a Weil sheaf on $G$. Every $g\in G(\Fq)$
determines a point $\bg$ on $\bG$ fixed by $\Frob{G}$ and
therefore an automorphism $\phi_{\bg}$ of $\gqcs{L}_{\bg}$. Set $\trFrob{(\gqcs{L},\phi)}(g) \ceq \trace(\phi_{\bg}, \gqcs{L}_{\bg})$,
and note that if $(\gqcs{L},\phi) \iso (\gqcs{L'},\phi')$ then $\trFrob{(\gqcs{L},\phi)} = \trFrob{(\gqcs{L'},\phi')}$.
 If $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf
then we will abbreviate $\trFrob{(\gqcs{L},\phi)}$ to $\trFrob{\qcs{L}}$.

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $g_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $g_\mathcal{L} = g_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
\QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $g_\qcs{L}$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.
It is easy to see that the function $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$ is a group homomorphism:
\[
g_{\qcs{L}\otimes \qcs{L'}} = g_{\qcs{L}} \, g_\qcs{L'}
\]
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.\todo{I still need to write down the details, for ourselves.}

@@

If $G$ is a connected and of finite type, then $\Hh^2(G(\bFq),\EEx)$ is trivial\todo{Reference?}
so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible;
 by contrast, if $G$ is etale then $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$ may be non-trivial so $G$ may admit invisible quasicharacters sheaves.
We will return to that case in Section~\ref{sec:etale}.

While neither $T_{G}$ nor $S_{G}$ are injective, their direct sum is.

\begin{proposition}
Let
\[
R_{G} : \QCiso{G} \to  \Hh^0(\Weil{},\Hh^2(G(\bFq),\EEx) \oplus \Hh^1(\Weil{}, \Hh^1(G(\bFq),\EEx))
\]
be the map formed from  . . .
Then $R_{G}$ is injective.
\end{proposition}

@@

\section{The function--sheaf dictionary for smooth, commutative group schemes}
\label{sec:snake}
